### 2024-12-06 README.Rmd
update README

# hatchR <a href="https://bmait101.github.io/hatchR/"><img src="man/figures/logo.png" align="right" height="139" alt="hatchR website" /></a>

<!-- badges: start -->
[![R-CMD-check](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**IN DEVELOPMENT**

**hatchR** is an R package for predicting fish developmental phenology using statistical models. It is intended to bridge the analytic gap of taking statistical models developed in hatchery settings (e.g., Beacham and Murray 1990) and applying them to real world temperature data sets using the effective value framework developed by Sparks et al. (2019).

@@

If you use **hatchR** in a formal publication or report, please cite it. Citing **hatchR** lets us devote more resources to it in the future. View the **hatchR** citation by running
```r
citation(package = "hatchR")
```

@@

``` r
# install.packages("pak")
pak::pak("bmait101/hatchR")
```

@@

1. Open a GitHub issue [link here](https://github.com/bmait101/hatchR/issues).
2. Email the hatchR support team ([Morgan Sparks](morgan.sparks@usda.gov), [Bryan Maitland](bryan.maitland@usda.gov))
3. Post on a support website like Stack Overflow or Cross Validated.

@@

The example `woody_island` dataset is included in **hatchR** and contains temperature data from Woody Island, Lake Iliamna, Alaska. It includes daily water temperature data from 1990-1992. We can use `plot_check_temp()` to visually check the data:

```{r}
p <- plot_check_temp(
  data = woody_island,
  dates = date, 
  temperature = temp_c
  )
p
```

Spawning in this system typically peaks around August 18 and hatching and emergence are done before the start of the following spawning season, so we can predict phenology within a subset of a year.

```{r}
p + 
  geom_rect(
    aes(
      xmin = ymd("1990-08-18"),  # spawn date
      xmax = ymd("1991-04-01"),  # approx phenology window end
      ymin = -10,  # lower bound
      ymax = 25),  # upper bound
    fill = "darkgreen",  # color rectangle dark green
    alpha = 0.01  # make rectangle semi-transparent
    ) 
```

@@

* dplyr: For data manipulation.
* ggplot2: For vignette visualizations.
* lubridate: For working with dates.
* tibble: For creating tibbles as output for various functions.
* rlang: For `.data` usage.
* stats: For various modeling functions (e.g., `confint()`).
* utils: For various utility functions.

@@

* knitr: For vignette building.
* readr: For reading in data.
* rmarkdown: For vignette building.
* testthat: For unit testing.

>>>>

# hatchR <a href="https://bmait101.github.io/hatchR/"><img src="man/figures/logo.png" alt="hatchR website" align="right" height="139"/></a>

<!-- badges: start -->

[![R-CMD-check](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

**hatchR** is an R package for predicting fish developmental phenology using statistical models. It is intended to bridge the analytic gap of taking statistical models developed in hatchery settings (*e.g.*, Beacham and Murray 1990) and applying them to real world temperature data sets using the effective value framework developed by Sparks et al. (2019).

@@

If you use **hatchR** in a formal publication or report, please cite it. Citing **hatchR** lets us devote more resources to it in the future. View the **hatchR** citation by running

```{r}
citation(package = "hatchR")
```

@@

```{r}
# install.packages("pak")
pak::pak("bmait101/hatchR")
```

@@

1.  Open a GitHub issue [link here](https://github.com/bmait101/hatchR/issues).
2.  Email the **hatchR** support team ([Morgan Sparks](morgan.sparks@usda.gov), [Bryan Maitland](bryan.maitland@usda.gov))
3.  Post on a support website like Stack Overflow or Cross Validated.

@@

The example `woody_island` dataset is included in **hatchR** and contains temperature data from Woody Island, Lake Iliamna, Alaska. It includes daily water temperature data from 1990-1992. We can use `plot_check_temp()` to visually check the data. Spawning in this system typically peaks around August 18 and hatching and emergence are done before the start of the following spawning season, so we can predict phenology within a subset of a year. Because `plot_check_temp()` returns a **ggplot2** object, we can add additional **ggplot2** elements to customize the plots.

```{r}
plot_check_temp(
  data = woody_island,
  dates = date, 
  temperature = temp_c
  ) +  # additional ggplot2 elements below this `+`
  geom_rect(
    aes(
      xmin = ymd("1990-08-18"),  # spawn date
      xmax = ymd("1991-04-01"),  # approx phenology window end
      ymin = -10,  # lower bound
      ymax = 25),  # upper bound
    fill = "grey",
    alpha = 0.01
    ) 
```

@@

-   dplyr: For data manipulation.
-   ggplot2: For vignette visualizations.
-   lubridate: For working with dates.
-   tibble: For creating tibbles as output for various functions.
-   rlang: For `.data` usage.
-   stats: For various modeling functions (e.g., `confint()`).
-   lifecycle: for function lifecycle badges.
-   utils: For various utility functions.

@@

-   knitr: For vignette building.
-   readr: For reading in data.
-   purrr: For iteration in vignettes.
-   rmarkdown: For vignette building.
-   nycflights13: For example data.
-   tidyr: For vignette building.
-   stringr: For vignette building.
-   ggridges: For vignette building.
-   patchwork: For vignette building.
-   cli: For better messages.
-   testthat: For unit testing.
